---
title: "STAT4420 Midterm"
author: "Alvin Xu"
date: today
format:
  pdf: default
  html:
    theme: cosmo
    toc: true
    code-fold: true
    embed-resources: true
execute: 
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
library(pacman)
p_load(tidyverse, gridExtra, forecast)
```

The data consists of the number of bikes $b_i$ and vehicles $v_i$ on ten
different streets. We will focus on modeling $y_i = b_i + v_i$, the
total amount of traffic (i.e. bikes + vehicles) on each street
(eg. $y_1 = 16 + 58 = 74$). Specifically, we consider the following
model for the total amount of traffic $y_i$ on street $i$:

$$\begin{aligned}
y_i &\sim \text{Poisson}(\theta_i) \\
\theta_i &\sim \text{Gamma}(\alpha, \beta)
\end{aligned}$$

# EDA
```{r}
df <- read.table("bicycles.txt", header = TRUE)
head(df)
```


# Question 1 (10 points)

Use Bayes Rule to derive the formula for the joint posterior
distribution $p(\alpha,\beta,\boldsymbol{\theta}|y)$ using a flat prior
distribution for $\alpha$ and $\beta,p(\alpha,\beta)\propto1$.

$$p(\alpha, \beta, \boldsymbol{\theta} | \boldsymbol{y}) \propto p(\boldsymbol{y} | \boldsymbol{\theta}) \cdot p(\boldsymbol{\theta} | \alpha, \beta) \cdot p(\alpha, \beta)$$
$$p(\boldsymbol{y} | \boldsymbol{\theta}) \propto \prod_{i=1}^{10} \theta_i^{y_i} e^{-\theta_i}$$

$$p(\boldsymbol{\theta} | \alpha, \beta) = \prod_{i=1}^{10} \frac{\beta^\alpha}{\Gamma(\alpha)} \theta_i^{\alpha-1} e^{-\beta \theta_i}$$

$$p(\alpha, \beta) \propto 1$$
Joint posterior distribution:
$$p(\alpha, \beta, \boldsymbol{\theta} | \boldsymbol{y}) \propto \left( \prod_{i=1}^{10} \theta_i^{y_i} e^{-\theta_i} \right) \cdot \left( \prod_{i=1}^{10} \frac{\beta^\alpha}{\Gamma(\alpha)} \theta_i^{\alpha-1} e^{-\beta \theta_i} \right) \cdot 1$$

Simplify:
$$p(\alpha, \beta, \boldsymbol{\theta} | \boldsymbol{y}) \propto \left( \frac{\beta^\alpha}{\Gamma(\alpha)} \right)^{10} \left( \prod_{i=1}^{10} \theta_i^{y_i + \alpha - 1} \right) e^{-(1+\beta)\sum_{i=1}^{10} \theta_i}$$

# Question 2 (10 points)

We want to use a Gibbs sampler to get samples from the joint posterior
distribution $p(\alpha,\beta,\theta|y)$. Derive the conditional
distribution $p(\theta_{i}|\alpha,\beta,y)$ for each $\theta_{i}$ given
all the other parameters. Derive the conditional distributions
$p(\alpha|\boldsymbol{\theta},\beta,y)$ and
$p(\beta|\boldsymbol{\theta},\alpha,y)$ for $\alpha$ and $\beta$ given all the other
parameters.

To find $p(\theta_i | \alpha, \beta, \boldsymbol{y})$, we should isolate the terms in the joint posterior that contain a specific $\theta_i$. We can ignore $\alpha$, $\beta$, and all other $\theta_{j \neq i}$ as they are treated as constants.$$p(\theta_i | \alpha, \beta, y_i) \propto \theta_i^{y_i + \alpha - 1} e^{-(1+\beta)\theta_i}$$

$$p(\theta_i | \alpha, \beta, y_i) \propto \theta_i^{y_i + \alpha - 1} e^{-(1+\beta)\theta_i}$$
The conditional distribution for each $\theta_{i}$ given all the other parameters is a Gamma distribution.

$$\theta_i | \alpha, \beta, y_i \sim \text{Gamma}(y_i + \alpha, 1 + \beta)$$

$$p(\beta | \boldsymbol{\theta}, \alpha, \boldsymbol{y}) \propto (\beta^\alpha)^{10} e^{-\beta \sum_{i=1}^{10} \theta_i}$$

$$p(\beta | \boldsymbol{\theta}, \alpha, \boldsymbol{y}) \propto \beta^{10\alpha} e^{-\beta \sum_{i=1}^{10} \theta_i}$$

$$p(\beta | \boldsymbol{\theta}, \alpha, \boldsymbol{y}) \propto \beta^{(10\alpha + 1) - 1} e^{-\beta \sum_{i=1}^{10} \theta_i}$$
The conditional distribution for $\beta$ given all the other parameters is:

$$\beta | \boldsymbol{\theta}, \alpha, \boldsymbol{y} \sim \text{Gamma}\left(10\alpha + 1, \sum_{i=1}^{10} \theta_i\right)$$

$$p(\alpha | \boldsymbol{\theta}, \beta, \boldsymbol{y}) \propto \left( \frac{\beta^\alpha}{\Gamma(\alpha)} \right)^{10} \prod_{i=1}^{10} \theta_i^\alpha$$

$$p(\alpha | \boldsymbol{\theta}, \beta, \boldsymbol{y}) \propto \frac{\beta^{10\alpha}}{\Gamma(\alpha)^{10}} \left( \prod_{i=1}^{10} \theta_i \right)^\alpha$$

$$p(\alpha | \boldsymbol{\theta}, \beta, \boldsymbol{y}) \propto \frac{ \left( \beta^{10} \prod_{i=1}^{10} \theta_i \right)^\alpha}{\Gamma(\alpha)^{10}}$$
We are unable to simplify the conditional distribution of $\alpha$ given all other parameters into a standard distribution, which is why we need to use the M-H algorithm.

# Question 3 (10 points)

The conditional distribution $p(\alpha|\boldsymbol{\theta},\beta,y)$ is
non-standard, so you will need to use a Metropolis (or
Metropolis-Hastings) step to obtain samples from
$p(\alpha|\boldsymbol{\theta},\beta,y)$. Choose a proposal distribution
for $\alpha$ and give the formula for the Metropolis (or Metropolis-Hastings)
ratio $r$ with your choice of proposal distribution.

Since $\alpha$ is a shape parameter for the Gamma distribution, it must be strictly positive ($\alpha > 0$). We can use a Normal distribution and reject negative values. The Normal distribution should be centered at $\alpha$.

$$\alpha^* \sim \text{Normal}(\alpha^{(t-1)}, \sigma_{\text{tune}}^2)$$

If $\alpha^* \leq 0$, when we will reject it. Because a Normal distribution is symmetric, the transition probabilities cancel out in the ratio: $q(\alpha^{(t-1)} | \alpha^*) = q(\alpha^* | \alpha^{(t-1)})$. Therefore, the Metropolis ratio $r$ simplifies to just the ratio of the target conditional posteriors.

$$r = \frac{p(\alpha^* | \boldsymbol{\theta}, \beta, \boldsymbol{y})}{p(\alpha^{(t-1)} | \boldsymbol{\theta}, \beta, \boldsymbol{y})} = \frac{ \frac{\beta^{10\alpha^*}}{\Gamma(\alpha^*)^{10}} \left( \prod_{i=1}^{10} \theta_i \right)^{\alpha^*} }{ \frac{\beta^{10\alpha^{(t-1)}}}{\Gamma(\alpha^{(t-1)})^{10}} \left( \prod_{i=1}^{10} \theta_i \right)^{\alpha^{(t-1)}} }$$

Simplify,
$$r = \left( \frac{\Gamma(\alpha^{(t-1)})}{\Gamma(\alpha^*)} \right)^{10} \beta^{10(\alpha^* - \alpha^{(t-1)})} \left( \prod_{i=1}^{10} \theta_i \right)^{\alpha^* - \alpha^{(t-1)}}$$
Our acceptance conditition is $\min(1, r)$.

# Question 4 (30 points)

Combine the Metropolis step from Question 3 with the conditional
distributions from Question 2 to form a Gibbs sampler, which iterates
between:

1.  sampling $\theta_{i}|\alpha,\beta,\boldsymbol{y}$ for each $i$
2.  sampling $\beta|\alpha,\boldsymbol{\theta},\boldsymbol{y}$
3.  sampling $\alpha|\beta,\boldsymbol{\theta},\boldsymbol{y}$

Use your Gibbs sampler to obtain 1000 independent posterior samples from
$p(\alpha,\beta,\boldsymbol{\theta}|\boldsymbol{y})$. Provide
convergence plots involving multiple chains run from different starting
points. How many burn-in iterations were needed to achieve convergence?
For the Metropolis sampling of
$\alpha|\beta,\boldsymbol{\theta},\boldsymbol{y}$, you should also give
the values of any tuning parameter that you used and the acceptance rate
you get with that value.

```{r}
# Read in data
row_idx <- 1

b_cols <- paste0("b", 1:10)
v_cols <- paste0("v", 1:10)

b <- as.numeric(df[row_idx, b_cols])
v <- as.numeric(df[row_idx, v_cols])

y <- b + v
n <- length(y)
```

```{r}
run_chain <- function(start_alpha, start_beta, tune_sd, num_iters) {
  # Setup storage (numeric() is slightly faster than rep(NA))
  alpha <- numeric(num_iters)
  beta  <- numeric(num_iters)
  theta <- matrix(NA_real_, nrow = num_iters, ncol = n)
  
  # Initialize
  alpha[1] <- start_alpha
  beta[1]  <- start_beta
  theta[1,] <- rgamma(n, shape = y + alpha[1], rate = 1 + beta[1])
  
  accept_count <- 0
  
  # Caching the log-gamma of alpha to avoid recalculating it every loop
  current_lgamma_alpha <- lgamma(alpha[1]) 
  
  for (t in 2:num_iters) {
    # Fetch previous values to avoid repeated indexing
    prev_alpha <- alpha[t-1]
    prev_beta <- beta[t-1]
    
    # Step 1: Sample theta (Store in a temporary variable first)
    current_theta <- rgamma(n, shape = y + prev_alpha, rate = 1 + prev_beta)
    theta[t,] <- current_theta
    
    # Step 2: Sample beta 
    current_beta <- rgamma(1, shape = n * prev_alpha + 1, rate = sum(current_theta))
    beta[t] <- current_beta
    
    # Step 3: Sample alpha 
    alpha_star <- rnorm(1, mean = prev_alpha, sd = tune_sd)
    
    if (alpha_star <= 0) {
      alpha[t] <- prev_alpha
    } else {
      # Calculate lgamma for the proposal
      lgamma_alpha_star <- lgamma(alpha_star)
      
      # Compute log-ratio using the cached lgamma and temporary variables
      log_r <- n * (current_lgamma_alpha - lgamma_alpha_star) +
               n * (alpha_star - prev_alpha) * log(current_beta) +
               (alpha_star - prev_alpha) * sum(log(current_theta))
      
      # Accept or reject
      if (log(runif(1)) < log_r) {
        alpha[t] <- alpha_star
        current_lgamma_alpha <- lgamma_alpha_star # Cache the new value for the next iteration!
        accept_count <- accept_count + 1
      } else {
        alpha[t] <- prev_alpha
      }
    }
  }
  
  accept_rate <- accept_count / (num_iters - 1)
  return(list(alpha = alpha, beta = beta, theta = theta, accept_rate = accept_rate))
}

# Run multiple chains with different starting values
num_iters <- 3000
tune_sd <- 3.8  # <- TUNING PARAMETER

set.seed(4420)
chain1 <- run_chain(start_alpha = 1, start_beta = 1, tune_sd = tune_sd, num_iters = num_iters)
chain2 <- run_chain(start_alpha = 10, start_beta = 5, tune_sd = tune_sd, num_iters = num_iters)

# Acceptance Rates
cat("Chain 1 Acceptance Rate:", round(chain1$accept_rate, 3), "\n")
cat("Chain 2 Acceptance Rate:", round(chain2$accept_rate, 3), "\n")

# 4. Convergence Plots
plot.frame <- data.frame(
  iterations = 1:num_iters,
  alpha_1 = chain1$alpha, alpha_2 = chain2$alpha,
  beta_1  = chain1$beta,  beta_2  = chain2$beta
)

plot_list <- vector("list", length = 2)

# Plot for Alpha
p1 <- ggplot(plot.frame) +
  geom_line(aes(x = iterations, y = alpha_1), col = "red", alpha = 0.7) +
  geom_line(aes(x = iterations, y = alpha_2), col = "blue", alpha = 0.7) +
  labs(title = "Trace Plot for Alpha", y = "Alpha", x = "Iteration")
plot_list[[1]] <- p1

# Plot for Beta
p2 <- ggplot(plot.frame) +
  geom_line(aes(x = iterations, y = beta_1), col = "red", alpha = 0.7) +
  geom_line(aes(x = iterations, y = beta_2), col = "blue", alpha = 0.7) +
  labs(title = "Trace Plot for Beta", y = "Beta", x = "Iteration")
plot_list[[2]] <- p2

grid.arrange(plot_list[[1]], plot_list[[2]], ncol = 1)
```
The goal of acceptance is usually around 50%, when the tuning parameter, SD was at 0.5, the acceptance values were around 75-80% for each of the chains, so I needed to modify it.

I increased the SD tuning parameter to 1.5 and got an acceptance rate of around 50% for each chain.

Apparently there is an optimal amount of acceptance: [https://arxiv.org/html/2408.06894v1](https://arxiv.org/html/2408.06894v1), so I will be tuning it even further.

I made the SD 3.8, and got an acceptance value of 0.239 and 0.232 for each chain.


# Question 5 (10 points)

Give autocorrelation plots for each parameter from the chains you ran in
Question 4 (after removing the burn-in iterations). How much thinning
was needed to obtain independent samples?

```{r}
set.seed(42
         )
# 1. Define Burn-in and Total Iterations, rerun chains
burn_in <- 500
num_iters <- 65500 # to ensure we end up with 1000 samples

chain1 <- run_chain(start_alpha = 1, start_beta = 1, tune_sd = 3.8, num_iters = num_iters)
chain2 <- run_chain(start_alpha = 10, start_beta = 5, tune_sd = 3.8, num_iters = num_iters)

# 2. Discard Burn-in and Combine Chains
alpha_post <- c(chain1$alpha[(burn_in + 1):num_iters], chain2$alpha[(burn_in + 1):num_iters])
beta_post  <- c(chain1$beta[(burn_in + 1):num_iters],  chain2$beta[(burn_in + 1):num_iters])

# Since theta is a matrix (n columns), we use rbind instead of c()
theta_post <- rbind(chain1$theta[(burn_in + 1):num_iters, ], 
                    chain2$theta[(burn_in + 1):num_iters, ])

total_post_samples <- length(alpha_post) # Should be 5000 (2500 from each chain)

# 3. Plot Autocorrelation for the un-thinned chains
p1 <- ggAcf(alpha_post, lag.max = 150) + labs(title = "ACF for Alpha")
p2 <- ggAcf(beta_post, lag.max = 150) + labs(title = "ACF for Beta")
grid.arrange(p1, p2, ncol = 1)
```

After thinning by taking every $130^{th}$ sample, the autocorrelation decreased to nearly 0.
```{r}
# 4. Thin the chains
thin_interval <- 130
thin_indices <- seq(from = thin_interval, to = total_post_samples, by = thin_interval)

alpha_final <- alpha_post[thin_indices]
beta_final  <- beta_post[thin_indices]
theta_final <- theta_post[thin_indices, ]

# 5. Verify the thinning worked
p3 <- ggAcf(alpha_final) + labs(title = "ACF for Alpha (Thinned)")
p4 <- ggAcf(beta_final) + labs(title = "ACF for Beta (Thinned)")
grid.arrange(p3, p4, ncol = 1)

# Number of independent samples
print(length(alpha_final))
```


# Question 6 (15 points)

Calculate the posterior means and 95% posterior intervals for a, and
each $\theta_{i}$ and provide these values in a table.

```{r}
# Calculate the mean and 95% CI for Alpha
alpha_mean <- mean(alpha_final)
alpha_ci <- quantile(alpha_final, probs = c(0.025, 0.975))

# Calculate the means and 95% CIs for all 10 Theta parameters
theta_means <- apply(theta_final, 2, mean)
theta_cis <- apply(theta_final, 2, quantile, probs = c(0.025, 0.975))

summary_table <- data.frame(
  Parameter = c("Alpha", paste0("Theta_", 1:10)),
  Mean = round(c(alpha_mean, theta_means), 3),
  `Lower_2.5` = round(c(alpha_ci[1], theta_cis[1, ]), 3),
  `Upper_97.5` = round(c(alpha_ci[2], theta_cis[2, ]), 3)
)

summary_table
```


# Question 7 (15 points)

Use your posterior samples to calculate a 95% predictive interval for
the total traffic on a new residential street with a bike lane.

```{r}
num_samples <- length(alpha_final)

theta_new <- rgamma(num_samples, shape = alpha_final, rate = beta_final)

y_new <- rpois(num_samples, lambda = theta_new)

predictive_interval <- quantile(y_new, probs = c(0.025, 0.975))

cat("95% Predictive Interval for total traffic on a new residential street with a bike lane:\n")
print(predictive_interval)
```

